Viscous Transport in Eroding Porous Media

### 2019-09-11 docs/transport/erosion_transport.tex
few changes in intro

Flow in porous media is often studied for a fixed geometry, but for many
physical processes, the geometry evolves with the flow. For example,
melting~\cite{}\todo[inline]{Natural convection solid/liquid phase
change in porous media}, dissolution~\cite{kan-zha-che-he2002},
deposition~\cite{joh-eli1995}, and biofilm
growth~\cite{tan-val-wer2015}.  Here, we focus on erosion which is
fluid-mechanical process that is prevalent in many geophysical and
hydrological applications.  For example, contaminant transport through
aquifers~\cite{cve-che-wen1996} and groundwater~\cite{dag1987,
kon-bre1978} is largely effected by the nature of the eroded porous
geometry.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that vary over several orders of
magnitude~\cite{all-hea-lab-rei2002} (see
Figure~\ref{fig:Eroding50vel}).  Moreover, the channelization results in
a geometry that is largely heterogeneous and anisotropic, and this
affects the transfer other other quantities such as a contaminant or
heat~\cite{nil-sto1990,
ree-sto1995}.  Others have studied transport in heterogeneous geometries
(for example, see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000,
hak-com-den2019, suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996,
leb-ded-dav-bou2007, knu-car2005, bel-sal-rin1992}). However, to the
best of our knowledge, a detailed study of transport in eroded
geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
fluid, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, qua-moo2018, par-izu2000}.  A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas or lattice Boltzmann
model~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.  Given the solution of the incompressible Stokes equations in
an eroded porous geometry, we investigate different different but
related metrics to
characterize transport.  We focus on three metrics: the tortuosity, the
anomalous dispersion, and the distribution of geometrical channels.

The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
diffusion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Tracer trajectories can be simulated
with a continuous time random walks~\cite{mon-wei1965,
den-cor-sch-ber2004, leb-den-car2008, ber-cor-den-sch2006} or with a
Fokker-Planck equation~\cite{ica-den2019}.  Instead, since there is no
stiff diffusive term in~\eqref{eqn:tracers}, we compute streamlines by
applying a high-order explicit Runge-Kutta time stepping method.

\todo[inline]{Bryan left off here}

@@

>>>>

Flow in porous media is often studied for a fixed geometry, but for many
physical processes, the geometry evolves with the flow. For example,
melting~\cite{}\todo[inline]{Natural convection solid/liquid phase
change in porous media}, dissolution~\cite{kan-zha-che-he2002},
deposition~\cite{joh-eli1995}, and biofilm
growth~\cite{tan-val-wer2015}.  Here, we focus on erosion which is
fluid-mechanical process that is prevalent in many geophysical and
hydrological applications.  For example, contaminant transport through
aquifers~\cite{cve-che-wen1996} and groundwater~\cite{dag1987,
kon-bre1978} is largely effected by the nature of the eroded porous
geometry.  Porous geometries that have undergone erosion have
certain characteristics.  For example, the geometry often has regions of
high porosity that connect the inlet to the outlet.  This channelization
results in velocity scales that vary over several orders of
magnitude~\cite{all-hea-lab-rei2002} (see
Figure~\ref{fig:Eroding50vel}).  Moreover, the channelization results in
a geometries that are largely heterogeneous and anisotropic, and this
affects the transfer other other quantities such as a contaminant or
heat~\cite{nil-sto1990, ree-sto1995}.  Others have studied transport in
heterogeneous geometries (for example,
see~\cite{mil-chr-imh-mcb-ped1998, ber-sch-sil2000, hak-com-den2019,
suo-liu-gan2019, cus-hu-den1995, cve-che-wen1996, leb-ded-dav-bou2007,
knu-car2005, bel-sal-rin1992}). However, to the best of our knowledge, a
detailed study of transport in eroded geometries has not been performed.

To describe the relevant hydrodynamics, we model two-dimensional
microscopic scales where the fluid is governed by the incompressible
Stokes equations (for example, see~\cite{hym-den-hag-kan2019,
den-ica-hid2018, bar-mar-vee-zha2018}).  To couple the erosion and the
fluid, the grains of the porous media are eroded at a rate that
is proportional to the hydrodynamic shear stress~\cite{wan-fel2004,
ris-moo-chi-she-zha2012, par-izu2000}.  Porous geometries
that have undergone erosion differ from standard porous geometry because
of large amounts of anisotropy and channelization. This work
addresses the effects that erosion has on the transport through a porous
media.  Our contributions are two-fold. First, starting with our
previous work~\cite{qua-moo2018}, we develop new numerical methods to
simulate erosion of geometries with low porosities and nearly-touching
grains. Second, we compute and analyze three different metrics of
transport throughout the erosion process: the tortuosity, the anomalous
dispersion, and the pore size distribution.

To calculate the tortuosity and dispersion
The first two of these metrics can be defined in terms of statistical
moments of trajectories.  Depending on the application, transport at the
microscale can be modelled as pure advection~\cite{dea-qua-bir-jua2018,
leb-ded-dav-bou2007, cve-che-wen1996, puy-gou-den2019}, and advection
diffusion equation~\cite{cus-hu-den1995, dag1987, den-ica-hid2018}, or
with a random walk~\cite{saf1959, bij-blu2006, ber-sch-sil2000}.  In
this work, we assume that the trajectories are governed purely by
advection (ie.~no diffusion).  Therefore, a trajectory $\ss(t)$
initialized at $\ss_0$ is governed by
\begin{align}
  \frac{d\ss}{dt} = \uu(\ss,t), \quad \ss(0) = \ss_0,
  \label{eqn:tracers}
\end{align}
where $\uu$ is the fluid velocity.  Tracer trajectories can be simulated
with a continuous time random walks~\cite{mon-wei1965,
den-cor-sch-ber2004, leb-den-car2008, ber-cor-den-sch2006} or with a
Fokker-Planck equation~\cite{ica-den2019}.  Instead, since there is no
stiff diffusive term in~\eqref{eqn:tracers}, we compute streamlines by
applying a high-order explicit Runge-Kutta time stepping method.

@@

A variety of
numerical methods can be applied to solve the incompressible Stokes
equations such as a lattice-gas or lattice Boltzmann
model~\cite{kop-kat-tim1996, dar-mcc1998}, finite
volume~\cite{suo-liu-gan2019, den-ica-hid2018, sie-ili-pri-riv-gua2019},
finite differences~\cite{leb-ded-dav-bou2007, knu-car2005}, and cellular
automata~\cite{rot1988}. In this work, we instead solve an equivalent
boundary integral equation (BIE) formulation of the incompressible
Stokes equation.  The BIE is discretized with a high-order quadrature
method that resolves nearly touching bodies with a near-singular
integration scheme~\cite{bar-wu-vee2015}.  The erosion dynamics are
simulated by applying a stable second-order time stepping Runge-Kutta
method to the {\thL} formulation~\cite{hou-low-she1994} of the eroding
bodies.
